---
title: "Martínez-Ugalde et al. 2026 - Diversity and Network Analysis"
description: This is an R notebook that contains all the commands used to perform diversity and network analysis presented in the "Microbial community networks reveal ecological consequences of soil management in Mexico City conservation landscape" manuscript. To run these commands you must load the functions provided in the "Martínez-Ugalde et al. 2026 - In house R functions" notebook.
---
##############################
1. Prepare an R env to run the analysis
##############################
1.1 Setwd
```{r}
setwd("~/Desktop/2026_papers/CDMX_networks/")
```

1.2 Load CRAN and GitHub packages
```{r}
cran_packages = c("tidyverse","devtools","dplyr","igraph",
                  "janitor","ggraph","patchwork","mclust",
                  "data.table","vegan","purrr","stringr",
                  "Biostrings","glue","ggrain","SpiecEasi",
                  "ggpubr","flextable","qiime2R","biomformat",
                  "Biostrings","sf","ggspatial","osmdata",
                  "ggnewscale","ggforce","ggrepel","scales",
                  "KEGGREST", "pulsar","huge", "pathcwork",
                  "cowplot", "colorspace")

sapply(c(cran_packages),
       function(pkg) suppressWarnings(suppressMessages(require(pkg,character.only=TRUE))))
```
##############################

##############################
2. Load tables, metadata and prepare data
##############################
```{r}
  # zOTU tables obtained from PipeCraft
bacteria_table = read.csv(file = "Feature_tables/16S_fixed.csv", sep = ",") |>
  dplyr::rename("Feature.ID" = "X.OTU.ID")

fungi_table = read.csv(file = "Feature_tables/ITS_fixed.csv", sep = ",") |>
  dplyr::rename("Feature.ID" = "X.OTU.ID")

metazoan_table = read.csv(file = "Feature_tables/COI_fixed.csv", sep = ",") |>
  dplyr::rename("Feature.ID" = "X.OTU.ID")

  # Taxonomy tables
bacteria_tax = read_qza(file = "Taxonomy/16S_taxonomy.qza")
bacteria_tax = bacteria_tax$data |>
  as.data.frame() |>
  separate(Taxon, into = c("Domain", "Phyla", "Class",
                           "Order","Family", "Genus",
                           "Species"), sep = ";") |>
  dplyr::select(-Confidence,-Species) |>
  filter(Domain != "d__Archaea")
  

fung_tax = read_qza(file = "Taxonomy/ITS_taxonomy.qza")
fung_tax = fung_tax$data |>
  as.data.frame() |>
  separate(Taxon, into = c("Domain", "Phyla", "Class",
                           "Order","Family", "Genus",
                           "Species"), sep = ";") |>
  dplyr::select(-Confidence,-Species) |>
  dplyr::filter(str_detect(Domain, "k__Fungi"))


metazoa_tax = read_qza(file = "Taxonomy/COI_taxonomy.qza")
metazoa_tax = metazoa_tax$data |>
  as.data.frame() |>
  separate(Taxon, into = c("Domain", "Phyla", "Class",
                           "Order","Family", "Genus",
                           "Species"), sep = ";") |>
  dplyr::select(-Confidence,-Species) |>
  dplyr::filter(!Phyla %in% c("p__Ascomycota_4890",
                              "p__Basidiomycota_5204",
                              "p__Chordata_7711"))

  # Metadata
metadata = read.csv(file = "Metadata/metadata.csv", sep = ",") |>
  dplyr::rename(sampleID = Sample) |>
  dplyr::rename(Latitude = Lat) |>
  dplyr::rename(Longitude = Long) |>
  dplyr::rename(Soil_system = Sistema) |>
  mutate(Soil_system = ifelse(Soil_system == "forestal",
                            "Conservation",Soil_system))|>
  mutate(Soil_system = ifelse(Soil_system == "agro",
                            "Agriculture",Soil_system))

  # Filter counts based on taxonomy 
bacteria_table = bacteria_table |>
  filter(Feature.ID %in% bacteria_tax$Feature.ID)

fungi_table = fungi_table |>
  filter(Feature.ID %in% fung_tax$Feature.ID)

metazoan_table = metazoan_table |>
  filter(Feature.ID %in% metazoa_tax$Feature.ID)
```
##############################

##############################
3. Perform diversity analysis for bacteria, fungi and metazoan datasets
##############################

3.1. Run rarefaction analysis
```{r}
  # Bacteria
  bacteria_rarefaction_summary = make_rarefaction_summary(count_table = bacteria_table,
                                                          feature_col = "Feature.ID",
                                                          rare_step = 100,
                                                          show_labels = FALSE)
  # Fungi
  fungi_rarefaction_summary = make_rarefaction_summary(count_table = fungi_table,
                                                          feature_col = "Feature.ID",
                                                          rare_step = 100,
                                                          show_labels = FALSE)
  # Metazoa
  metazoa_rarefaction_summary = make_rarefaction_summary(count_table = metazoan_table,
                                                          feature_col = "Feature.ID",
                                                          rare_step = 100,
                                                          show_labels = FALSE)
```

3.2 Visualization of rarefaction analysis
```{r}
  # Bacteria
  bacteria_rare_curve = bacteria_rarefaction_summary$rare_curve
  bacteria_retention_plot = bacteria_rarefaction_summary$retention_plot
  
  # Fungi
  fungi_rare_curve = fungi_rarefaction_summary$rare_curve
  fungi_retention_plot = fungi_rarefaction_summary$retention_plot
  
  # Metazoa
  metazoa_rare_curve = metazoa_rarefaction_summary$rare_curve
  metazoa_retention_plot = metazoa_rarefaction_summary$retention_plot
```

3.3 Calcualte diversity metrics
3.3.1 Select rarefaction treshold
```{r}
  # Tresholds selected preserve 90% of the samples based on rarefaction summary from 3.2
  bacteria_rare_tresh  = 10000
  fungi_rare_tresh = 41500
  metazoa_rare_tresh = 29000
```

3.3.2 Run rarefaction
```{r}
  # Bacteria
  bacteria_rare = rarefy_table(count_table = bacteria_table,
                                depth = bacteria_rare_tresh,
                                feature_col = "Feature.ID")
  # Fungi
  fungi_rare = rarefy_table(count_table = fungi_table,
                                depth = fungi_rare_tresh,
                                feature_col = "Feature.ID")
  # Metazoa
  metazoa_rare = rarefy_table(count_table = metazoan_table,
                                depth = metazoa_rare_tresh,
                                feature_col = "Feature.ID")
```

3.3.3 Calculation of alpha diversity metrics
```{r}
  # Alpha calculation
  bacteria_alpha = cal_alpha_div(bacteria_rare)
  fungi_alpha = cal_alpha_div(fungi_rare)
  metazoa_alpha = cal_alpha_div(metazoa_rare)
  
  # Join alpha results with metadata
  bacteria_alpha = bacteria_alpha |>
    left_join(metadata, by = "sampleID")
  
  fungi_alpha = fungi_alpha |>
    left_join(metadata, by = "sampleID")
  
  metazoa_alpha = metazoa_alpha |>
    left_join(metadata, by = "sampleID")
```

3.3.4 Calculation of beta diversity metrics
```{r}
  bacteria_beta = cal_beta_div(asv_table = bacteria_rare,
                               non_numeric_cols = "Feature.ID")

  fungi_beta = cal_beta_div(asv_table = fungi_rare, 
                            non_numeric_cols = "Feature.ID")
  
  metazoa_beta = cal_beta_div(asv_table = metazoa_rare, 
                              non_numeric_cols = "Feature.ID")
```

3.3.5 Statistical comparisons of alpha diversity
```{r}
  alpha_comp_formula = list(Shannon ~ Soil_system,
                            Richness ~ Soil_system,
                            BergerParker ~ Soil_system)
  
  bact_div_comp = compare_alpha_diversity(formulas = alpha_comp_formula, 
                                          data = bacteria_alpha)
  
  fung_div_comp = compare_alpha_diversity(formulas = alpha_comp_formula, 
                                          data = fungi_alpha)
  
  meta_div_comp = compare_alpha_diversity(formulas = alpha_comp_formula, 
                                          data = metazoa_alpha)
```

3.3.6 Statistical comparisons of beta diversity (PCoA and PERMANOVA)
```{r Beta statitistics}
  # Perform PCoA
  bacteria_pcoa = perform_pcoa(dist_matrix = bacteria_beta$Bray_Curtis)
  fungi_pcoa = perform_pcoa(dist_matrix = fungi_beta$Bray_Curtis)
  metazoa_pcoa = perform_pcoa(dist_matrix = metazoa_beta$Bray_Curtis)
  
  # PERMANOVA
  beta_formula= c("Soil_system")
  
  bacteria_permanova = perform_permanova(matrix_list = bacteria_beta$Bray_Curtis,
                                        metadata = bacteria_alpha,
                                        variables = beta_formula,
                                        permutations = 999)
  
  fungi_permanova = perform_permanova(matrix_list = fungi_beta$Bray_Curtis,
                                        metadata = fungi_alpha,
                                        variables = beta_formula,
                                        permutations = 999)
  
  metazoa_permanova = perform_permanova(matrix_list = metazoa_beta$Bray_Curtis,
                                        metadata = metazoa_alpha,
                                        variables = beta_formula,
                                        permutations = 999)
```

##############################

##############################
4. Visualization of diversity metrics
##############################
4.1 Generate alpha diversity plots
```{r}
  cdmx_cols = c("#D50B56","#D3DE00")
  
  # Bacteria
  shan_bact = plot_alpha(data = bacteria_alpha, x_var = "Soil_system",
                         y_var = "Shannon", colors = cdmx_cols,
                         title = "A", guide_title = "Management type")
  
  rich_bact = plot_alpha(data = bacteria_alpha, x_var = "Soil_system",
                         y_var = "Richness", colors = cdmx_cols,
                         title = "B")
  
  berg_bact = plot_alpha(data = bacteria_alpha, x_var = "Soil_system",
                         y_var = "BergerParker", colors = cdmx_cols,
                         title = "C")
  
  # Fungi
  shan_fung = plot_alpha(data = fungi_alpha, x_var = "Soil_system",
                         y_var = "Shannon", colors = cdmx_cols,
                         title = "D")
  
  rich_fung = plot_alpha(data = fungi_alpha, x_var = "Soil_system",
                         y_var = "Richness", colors = cdmx_cols,
                         title = "E")
  
  berg_fung = plot_alpha(data = fungi_alpha, x_var = "Soil_system",
                         y_var = "BergerParker", colors = cdmx_cols,
                         title = "F")
  
  # Metazoa
  shan_meta = plot_alpha(data = metazoa_alpha, x_var = "Soil_system",
                         y_var = "Shannon", colors = cdmx_cols,
                         title = "G")
  
  rich_meta = plot_alpha(data = metazoa_alpha, x_var = "Soil_system",
                         y_var = "Richness", colors = cdmx_cols,
                         title = "H")
  
  berg_meta = plot_alpha(data = metazoa_alpha, x_var = "Soil_system",
                         y_var = "BergerParker", colors = cdmx_cols,
                         title = "I")
```

4.2 Generate beta diversity plots
```{r}
  bact_pcoa_plot = plot_pcoa(pcoa_positions = bacteria_pcoa$positions,
                             metadata = metadata,
                             group_var = "Soil_system",
                             explained = bacteria_pcoa$explained,
                             title = "A", guide_title = "Management type",
                             colors = cdmx_cols)
  
  fung_pcoa_plot = plot_pcoa(pcoa_positions = fungi_pcoa$positions,
                             metadata = metadata,
                             group_var = "Soil_system",
                             explained = fungi_pcoa$explained,
                             title = "B", guide_title = "Management type",
                             colors = cdmx_cols)
  
  meta_pcoa_plot = plot_pcoa(pcoa_positions = metazoa_pcoa$positions,
                             metadata = metadata,
                             group_var = "Soil_system",
                             explained = metazoa_pcoa$explained,
                             title = "C", guide_title = "Management type",
                             colors = cdmx_cols)
```
##############################

##############################
5. Taxonomic profiles per microbial group
##############################
5.1 Clean taxonomy tables
```{r}
  bacteria_taxonomy_clean = clean_taxonomy_propagate(bacteria_tax)
  fungi_taxonomy_clean = clean_taxonomy_propagate(fung_tax)
  metazoa_taxonomy_clean = clean_taxonomy_propagate(metazoa_tax)
  
  metazoa_taxonomy_clean = metazoa_taxonomy_clean |>
    mutate(across(-Feature.ID, ~ gsub("_[0-9]+$", "", .)))
```

5.2 Compute taxonomic table summaries
```{r}
  # Bacteria abund summary
  bact_abund_summary = compute_tax_abundance_table(
    bacteria_table,
    bacteria_taxonomy_clean,
    metadata,
    tax_level = "Order",
    group_vars = c("Soil_system"))
  
  # Fungi abund summary
  fung_abund_summary = compute_tax_abundance_table(
    fungi_table,
    fungi_taxonomy_clean,
    metadata,
    tax_level = "Order",
    group_vars = c("Soil_system"))
  
  # Metazoa abund summary
  meta_abund_summary = compute_tax_abundance_table(
    metazoan_table,
    metazoa_taxonomy_clean,
    metadata,
    tax_level = "Order",
    group_vars = c("Soil_system"))
```

5.3 Generate relative abundance taxonomy barplots
```{r Taxonomy barplots}
  # Bacteria
  bacteria_taxa_barplot = make_rel_abundance_barplot(asv_table = bacteria_table,
                                                     taxonomy = bacteria_taxonomy_clean,
                                                     metadata = metadata,
                                                     tax_level = "Order",
                                                     group_vars = "Soil_system",
                                                     min_tax_rank = "Class",
                                                     low_freq_threshold = 0.02,
                                                     custom_colors = bacteria_cols,
                                                     blank_axis_titles = TRUE,
                                                     plot_title = "B")


  # Fungi
  fungi_taxa_barplot = make_rel_abundance_barplot(asv_table = fungi_table,
                                                       taxonomy = fungi_taxonomy_clean,
                                                       metadata = metadata,
                                                       tax_level = "Order",
                                                       group_vars = "Soil_system",
                                                       min_tax_rank = "Class",
                                                       low_freq_threshold = 0.02,
                                                       custom_colors = fungi_cols,
                                                       blank_axis_titles = TRUE,
                                                       plot_title = "C")
  
  # Metazoan
  metazoan_taxa_barplot = make_rel_abundance_barplot(asv_table = metazoan_table,
                                                       taxonomy = metazoa_taxonomy_clean,
                                                       metadata = metadata,
                                                       tax_level = "Order",
                                                       group_vars = "Soil_system",
                                                       min_tax_rank = "Class",
                                                       low_freq_threshold = 0.02,
                                                       custom_colors = metazoa_cols,
                                                       blank_axis_titles = TRUE,
                                                       plot_title = "D")
```
##############################

##############################
6. Create a map showing conservation soil and sampling locations (Figure1)
##############################
6.1 Map creation (shape files were downloaded from INEGI portal and are provided as shp files on the github repository
```{r}
  # Load data
  cons_soil_shp = st_read("Map_info/Conservation_soil/suelo-de-conservacion.shp")
  mexico_map = st_read("Map_info/Mexico_map/dest_2010cw.shp")
  human_sets = st_read("Map_info/Mexico_human_set/09_ciudaddemexico/conjunto_de_datos/09as.shp")

  # Get data for mexico city
  cdmx_map = mexico_map |>
    filter(ENTIDAD == "DISTRITO FEDERAL") 
  
  # Transform shp files to the same coordinate measure
  cdmx_map = st_transform(cdmx_map, crs = 4326)
  cons_soil_shp = st_transform(cons_soil_shp, crs = 4326)

  # Convert conservation poligons 
  cons_union = cons_soil_shp |> 
    st_union()

  # Set no conservation area
  non_cons = st_difference(cdmx_map, cons_union)
  
  # Get data about human settlings 
  human_sets = human_sets |> 
    mutate(tipo = "Human setting") |>
    dplyr::rename(Soil_type = tipo)

  # Data points
  points_sf = metadata |>
    dplyr::select(sampleID, Latitude, Longitude, Soil_system) |>
    st_as_sf(coords = c("Latitude", "Longitude"), crs = 4326)
    
  points_sf = st_transform(points_sf, st_crs(cdmx_map))

  # Plot the map 
  samp_map = ggplot() +
  geom_sf(data = cdmx_map, fill = NA, color = "black", size = 0.5) +
  geom_sf(data = non_cons, fill = "#6E6E6E", color = NA, alpha = 0.3) +
  geom_sf(data = cons_union, aes(fill = "Conservation soil"),
          color = NA, alpha = 0.7) +
  geom_sf(data = human_sets, aes(fill = "Urban soil"),
          color = NA, alpha = 0.3) +
  scale_fill_manual(name   = "Land use",
    values = c("Conservation soil" = "#4C624E","Urban soil" = "#6E6E6E"),
    guide = guide_legend(ncol = 1, override.aes = list(shape = 22, size = 3, 
                                             color  = NA, stroke = 0))) +
  new_scale_fill() +
  geom_sf(data = points_sf, aes(fill = Soil_system), shape = 21,
          color = "#2D3740FF", size = 3, stroke = 0.6) +
  scale_fill_manual(name   = "Soil management",values = c("#D50B56", "#D3DE00"),
                    guide  = guide_legend(ncol = 1, 
                                          override.aes = list(shape = 21, size = 3))) +
    labs(title = "A") +
    theme_minimal()+
    theme(plot.title.position = "right",
          plot.title = element_text(face = "bold", size = 18))
 
```
##############################

##############################
7. Prepare data for network analysis
##############################
7.1 Collapse count tables at genus level
```{r}
  # Bacteria 
  bacteria_table_genus = collapse_by_taxa(asv_table =  bacteria_table,
                                          taxonomy_table = bacteria_taxonomy_clean,
                                          level = "Genus")
  # Fungi
  fungi_table_genus = collapse_by_taxa(asv_table =  fungi_table,
                                       taxonomy_table = fungi_taxonomy_clean,
                                       level = "Genus") 
  # Metazoa
  metazoa_table_genus = collapse_by_taxa(asv_table =  metazoan_table,
                                         taxonomy_table = metazoa_taxonomy_clean,
                                         level = "Genus")
```

7.2 Split count tables by soil management type and filtering low abundant and prevalent genus
```{r}
  # Bacteria
  bacteria_network_matrices = split_management(counts_df = bacteria_table_genus,
                                               metadata = metadata,
                                               min_total = 100,
                                               min_prev = 0.1)
  # Fungi
  fungi_network_matrices = split_management(counts_df = fungi_table_genus,
                                            metadata = metadata,
                                            min_total = 100,
                                            min_prev = 0.1)
  # Metazoa
  metazoa_network_matrices = split_management(counts_df = metazoa_table_genus,
                                              metadata = metadata,
                                              min_total = 100,
                                              min_prev = 0.1)
```

7.3 Generate a combined dataset for multi-kingdom analysis
```{r}
 # Create a multikingdom dataset 
  bacteria_conservation = bacteria_network_matrices$Conservation
  fungi_conservation = fungi_network_matrices$Conservation

  bacteria_agriculture = bacteria_network_matrices$Agriculture
  fungi_agriculture = fungi_network_matrices$Agriculture

    # Verify row order before cbind
   identical(rownames(bacteria_conservation), rownames(fungi_conservation)) &&
     identical(rownames(bacteria_conservation), rownames(metazoa_conservation))
   
   identical(rownames(bacteria_agriculture), rownames(fungi_agriculture)) &&
     identical(rownames(bacteria_agriculture), rownames(metazoa_agriculture))
   
    # Obtain merged matrices
    multi_kingdom_conservation = cbind(bacteria_conservation,
                                       fungi_conservation)
    
    multi_kingdom_agriculture = cbind(bacteria_agriculture,
                                      fungi_agriculture)
```
##############################

##############################
8. Inferring Microbial co-occurrence networks using spiec-easi
##############################
8.1 Infer networks
```{r}
  # Bacteria netwokrs
  bact_cons_net = infer_network(bacteria_network_matrices$Conservation,
                                nlambda = 60,
                                lambda.min.ratio = 1e-3,
                                pulsar.n = 400,
                                pulsar.subsample.ratio = 0.8,
                                pulsar.thresh = 0.1,
                                seed = 1502)
  
  bact_agri_net = infer_network(bacteria_network_matrices$Agriculture,
                                nlambda = 60,
                                lambda.min.ratio = 1e-3,
                                pulsar.n = 400,
                                pulsar.subsample.ratio = 0.8,
                                pulsar.thresh = 0.1,
                                seed = 1502)

  # Fungi netwokrs
  fung_cons_net = infer_network(fungi_network_matrices$Conservation,
                                nlambda = 60,
                                lambda.min.ratio = 1e-3,
                                pulsar.n = 400,
                                pulsar.subsample.ratio = 0.8,
                                pulsar.thresh = 0.1,
                                seed = 1502)
  
  fung_agri_net = infer_network(fungi_network_matrices$Agriculture,
                                nlambda = 60,
                                lambda.min.ratio = 1e-3,
                                pulsar.n = 400,
                                pulsar.subsample.ratio = 0.8,
                                pulsar.thresh = 0.1,
                                seed = 1502)

  # Metazoo netwokrs
  meta_cons_net = infer_network(metazoa_network_matrices$Conservation,
                                nlambda = 60,
                                lambda.min.ratio = 1e-3,
                                pulsar.n = 400,
                                pulsar.subsample.ratio = 0.8,
                                pulsar.thresh = 0.1,
                                seed = 1502)
  
  meta_agri_net = infer_network(metazoa_network_matrices$Agriculture,
                                nlambda = 60,
                                lambda.min.ratio = 1e-3,
                                pulsar.n = 400,
                                pulsar.subsample.ratio = 0.8,
                                pulsar.thresh = 0.1,
                                seed = 1502)

 
  # Multi-kingdom networks
  multiK_cons_net = infer_network(multi_kingdom_conservation,
                                  nlambda = 80,
                                  lambda.min.ratio = 1e-4,
                                  pulsar.n = 400,
                                  pulsar.subsample.ratio = 0.9,
                                  pulsar.thresh = 0.1,
                                  seed = 1502)
  
  multikK_agri_net = infer_network(multi_kingdom_agriculture,
                                   nlambda = 80,
                                   lambda.min.ratio = 1e-4,
                                   pulsar.n = 400,
                                   pulsar.subsample.ratio = 0.9,
                                   pulsar.thresh = 0.1,
                                   seed = 1502)
```

8.2 Detecting modules
```{r}
  # Bacteria
  bact_cons_modules = detect_modules(bact_cons_net,comm_method = "louvain")
  bact_agri_modules = detect_modules(bact_agri_net,comm_method = "louvain")
  
  # Fungi
  fung_cons_modules = detect_modules(fung_cons_net,comm_method = "louvain")
  fung_agri_modules = detect_modules(fung_agri_net,comm_method = "louvain")
  
  # Metazoa
  meta_cons_modules = detect_modules(meta_cons_net,comm_method = "louvain")
  meta_agri_modules = detect_modules(meta_agri_net,comm_method = "louvain")
  
  # Multi-kingdom
  multiK_cons_modules = detect_modules(multiK_cons_net,comm_method = "louvain")
  multiK_agri_modules = detect_modules(multikK_agri_net,comm_method = "louvain")
```

8.3 Compute network metrics
```{r}
  # Bacteria metrics
  bact_cons_metrics = compute_metrics(bact_cons_net, 
                                      bact_cons_modules$comm,
                                      bact_cons_modules$membership,
                                      group_label = "Bacteria_Cons")
  
  bact_agri_metrics = compute_metrics(bact_agri_net, 
                                      bact_agri_modules$comm,
                                      bact_agri_modules$membership,
                                      group_label = "Bacteria_Agri")
  # Fungi metrics
  fung_cons_metrics = compute_metrics(fung_cons_net, 
                                      fung_cons_modules$comm,
                                      fung_cons_modules$membership,
                                      group_label = "Fungi_Cons")
  
  fung_agri_metrics = compute_metrics(fung_agri_net, 
                                      fung_agri_modules$comm,
                                      fung_agri_modules$membership,
                                      group_label = "Fungi_Agri")
  
  # Multi-kingdom
  multik_cons_metrics = compute_metrics(multiK_cons_net, 
                                         multiK_cons_modules$comm,
                                         multiK_cons_modules$membership,
                                         group_label = "MicrEuk_Cons")
  
  multik_agri_metrics = compute_metrics(multikK_agri_net, 
                                         multiK_agri_modules$comm,
                                         multiK_agri_modules$membership,
                                         group_label = "MicrEuk_Agri")
```

8.4 Select threshold to plot node degree based on the 3rd quantile distribution
```{r}
  # Bacteria
  # Quantile distribution node degree
  quantile(igraph::degree(bact_agri_net))
  quantile(igraph::degree(bact_cons_net))
  
  # Quantile distribution edge weight
  quantile(abs(E(bact_agri_net)$weight), probs = 0.90)
  quantile(abs(E(bact_cons_net)$weight), probs = 0.90)

  # Fungi
  # Quantile distribution node degree
  quantile(igraph::degree(fung_agri_net))
  quantile(igraph::degree(fung_cons_net))

  # Quantile distribution edge weight
  quantile(abs(E(fung_agri_net)$weight), probs = 0.90)
  quantile(abs(E(fung_cons_net)$weight), probs = 0.90)
```


8.5 Set network palette that preserve taxonomic identity and draw per group/management network plots
```{r}

  bacteria_palette = c(
    "d__Bacteria" = "#FFE4C4", 
    "p__Actinomycetota" = "#FF7F24",
    "p__Armatimonadota" = "#CDB79E",
    "p__Acidobacteriota" = "#D2691E",
    "p__Bacillota" = "#8B7D6B",
    "p__Bacteroidota" = "#CD661D",
    "p__Chlamydiota" = "#8B7355",
    "p__Chloroflexota" = "#8B4513",
    "p__Cyanobacteriota" = "#EEAD0E",
    "p__Entotheonellaeota" = "#FF7F50",
    "p__Fibrobacterota" = "#CDC8B1",
    "p__GAL15" = "#CD5B45",
    "p__Gemmatimonadota" = "#8B8878",
    "p__Latescibacterota" = "#8B3E2F",
    "p__Methylomirabilota" = "#B8860B",
    "p__Myxococcota" = "#EE7600",
    "p__Nitrospirota" = "#EE5C42",
    "p__Patescibacteria" = "#CD6600",
    "p__Planctomycetota" = "#8B6508",
    "p__Pseudomonadota" = "#8B4500",
    "p__RCP2−54" = "#EEDC82",
    "p__Thermodesulfobacteriota" = "#EE9572",
    "p__Verrucomicrobiota" = "#CDBE70",
    "p__WS2" = "#8B5742"
    )

  fungi_palette = c(
    "p__Ascomycota" = "#00BFFF",
    "p__Aphelidiomycota" = "#EE6AA7",
    "p__Basidiomycota" = "#009ACD",
    "p__Blastocladiomycota" = "#CD6090",
    "p__Calcarisporiellomycota" = "#00688B",
    "p__Chytridiomycota" = "#8B3A62",
    "p__Glomeromycota" = "#1C86EE",
    "p__Kickxellomycota" = "#EEA2AD",
    "p__Monoblepharomycota" = "#104E8B",
    "p__Mortierellomycota" = "#CD8C95",
    "p__Mucoromycota" = "#BA55D3",
    "p__Olpidiomycota" = "#8B5F65",
    "p__Zoopagomycota" = "#CD2990"
    )
  
  
  multik_palette = c(
    "d__Bacteria" = "#FFE4C4", 
    "p__Actinomycetota" = "#FF7F24",
    "p__Armatimonadota" = "#CDB79E",
    "p__Acidobacteriota" = "#D2691E",
    "p__Bacillota" = "#8B7D6B",
    "p__Bacteroidota" = "#CD661D",
    "p__Chlamydiota" = "#8B7355",
    "p__Chloroflexota" = "#8B4513",
    "p__Cyanobacteriota" = "#EEAD0E",
    "p__Entotheonellaeota" = "#FF7F50",
    "p__Fibrobacterota" = "#CDC8B1",
    "p__GAL15" = "#CD5B45",
    "p__Gemmatimonadota" = "#8B8878",
    "p__Latescibacterota" = "#8B3E2F",
    "p__Methylomirabilota" = "#B8860B",
    "p__Myxococcota" = "#EE7600",
    "p__Nitrospirota" = "#EE5C42",
    "p__Patescibacteria" = "#CD6600",
    "p__Planctomycetota" = "#8B6508",
    "p__Pseudomonadota" = "#8B4500",
    "p__RCP2−54" = "#EEDC82",
    "p__Thermodesulfobacteriota" = "#EE9572",
    "p__Verrucomicrobiota" = "#CDBE70",
    "p__WS2" = "#8B5742",
    "p__Ascomycota" = "#00BFFF",
    "p__Aphelidiomycota" = "#EE6AA7",
    "p__Basidiomycota" = "#009ACD",
    "p__Blastocladiomycota" = "#CD6090",
    "p__Calcarisporiellomycota" = "#00688B",
    "p__Chytridiomycota" = "#8B3A62",
    "p__Glomeromycota" = "#1C86EE",
    "p__Kickxellomycota" = "#EEA2AD",
    "p__Monoblepharomycota" = "#104E8B",
    "p__Mortierellomycota" = "#CD8C95",
    "p__Mucoromycota" = "#BA55D3",
    "p__Olpidiomycota" = "#8B5F65",
    "p__Zoopagomycota" = "#CD2990"
  )
  
  # Bacteria
  bact_agri_plot = plot_network_comparison(
    graph = bact_agri_net,
    membership = bact_agri_modules$membership,
    taxonomy = bacteria_taxonomy_clean,
    highlight_degree = 27,
    tax_rank = "Phyla",
    color_by_tax = TRUE,
    layout_algo = "graphopt", 
    fixed_node_size = 3,
    show_labels = FALSE,
    node_palette = bacteria_palette,        
    p3_node_color = "#B0BEC5",
    seed = 1502,
    min_abs_weight = 0.200,
    p1_name = "A",
    p2_name = "A",
    p3_name = "A"
  )

  bact_cons_plot = plot_network_comparison(
    graph = bact_cons_net,
    membership = bact_cons_modules$membership,
    taxonomy = bacteria_taxonomy_clean,
    highlight_degree = 28,
    tax_rank = "Phyla",
    color_by_tax = TRUE,
    layout_algo = "graphopt",  
    fixed_node_size = 3,
    show_labels = FALSE,
    node_palette = bacteria_palette,        
    p3_node_color = "#B0BEC5",
    seed = 1502,
    min_abs_weight = 0.224,
    p1_name = "B",
    p2_name = "B",
    p3_name = "B"
  )
  
  # Fungi
  fung_agri_plot = plot_network_comparison(
    graph = fung_agri_net,
    membership = fung_agri_modules$membership,
    taxonomy = fungi_taxonomy_clean,
    highlight_degree = 31,
    tax_rank = "Phyla",
    color_by_tax = TRUE,
    layout_algo = "graphopt", 
    fixed_node_size = 3,
    show_labels = FALSE,
    node_palette = fungi_palette,        
    p3_node_color = "#B0BEC5",
    seed = 1502,
    min_abs_weight = 0.211,
    p1_name = "C",
    p2_name = "C",
    p3_name = "C"
  )
  
  fung_cons_plot = plot_network_comparison(
    graph = fung_cons_net,
    membership = fung_cons_modules$membership,
    taxonomy = fungi_taxonomy_clean,
    highlight_degree = 33,
    tax_rank = "Phyla",
    color_by_tax = TRUE,
    layout_algo = "graphopt",  
    fixed_node_size = 3,
    show_labels = FALSE,
    node_palette = fungi_palette,        
    p3_node_color = "#B0BEC5",
    seed = 1502,
    min_abs_weight = 0.213,
    p1_name = "D",
    p2_name = "D",
    p3_name = "D"
  )
```
##############################

##############################
9. Classify node topological propierties using the Zi-Pi framework 
##############################
9.1 Calculate Zi-Pi roles
```{r}
  # Bacteria  Zi-Pi 
  bact_cons_zipi = compute_zipi_roles(bact_cons_net, 
                                      bact_cons_modules$membership)
  bact_agri_zipi = compute_zipi_roles(bact_agri_net, 
                                      bact_agri_modules$membership)
  
  # Fungi  Zi-Pi 
  fung_cons_zipi = compute_zipi_roles(fung_cons_net,
                                      fung_cons_modules$membership)
  fung_agri_zipi = compute_zipi_roles(fung_agri_net, 
                                      fung_agri_modules$membership)
  
  # Metazoa  Zi-Pi 
  meta_cons_zipi = compute_zipi_roles(meta_cons_net,
                                      meta_cons_modules$membership)
  meta_agri_zipi = compute_zipi_roles(meta_agri_net,
                                      meta_agri_modules$membership)

  # Multi-K  Zi-Pi 
  multiK_cons_zipi = compute_zipi_roles(multiK_cons_net,
                                        multiK_cons_modules$membership)
  multiK_agri_zipi = compute_zipi_roles(multikK_agri_net,
                                        multiK_agri_modules$membership)
```

9.2 Plot Zi-Pi results
```{r}
  # Bacteria
  bact_cons_zipi_plot = plot_zipi(bact_cons_zipi, label_nodes = TRUE,
                                  plot_title = "B")
  bact_agri_zipi_plot = plot_zipi(bact_agri_zipi, label_nodes = TRUE,
                                  plot_title = "A")
  
  # Fungi
  fung_cons_zipi_plot = plot_zipi(fung_cons_zipi, label_nodes = TRUE, 
                                  plot_title = "D")
  fung_agri_zipi_plot = plot_zipi(fung_agri_zipi, label_nodes = TRUE,
                                  plot_title = "C")

 # Multi-K
  multiK_cons_zipi_plot = plot_zipi(multiK_cons_zipi, label_nodes = TRUE,
                                    plot_title = "B")
  multiK_agri_zipi_plot = plot_zipi(multiK_agri_zipi, label_nodes = TRUE,
                                    plot_title = "A")
```
##############################

##############################
10. Perform a descriptive analysis of the multi-kingdom networks
##############################
10.1 Draw multikindom plot
```{r}
  taxonomy_all = rbind(bacteria_taxonomy_clean,
                       fungi_taxonomy_clean)

  taxonomy_all$Domain = gsub("^d__", "", taxonomy_all$Domain)
  taxonomy_all$Domain = gsub("^k__", "", taxonomy_all$Domain)


  multik_cons_plot = plot_network_comparison(
      graph = multiK_cons_net,
      membership = multiK_cons_modules$membership,
      taxonomy = taxonomy_all,
      color_by_tax = TRUE,
      tax_rank = "Phyla",
      highlight_degree = 59,
      layout_algo = "graphopt",  
      fixed_node_size = 2,
      show_labels = FALSE,
      node_palette = multik_palette,        
      p3_node_color = "#B0BEC5",
      min_abs_weight = 0.223,
      seed = 1502,
      p1_name = "B",
      p2_name = "B",
      p3_name = "B"
    )


  multik_agri_plot = plot_network_comparison(
      graph = multikK_agri_net,
      membership = multiK_agri_modules$membership,
      taxonomy = taxonomy_all,
      color_by_tax = TRUE,
      tax_rank = "Phyla",
      highlight_degree = 61,
      layout_algo = "graphopt",  
      fixed_node_size = 2,
      show_labels = FALSE,
      node_palette = multik_palette,        
      p3_node_color = "#B0BEC5",
      min_abs_weight = 0.214,
      seed = 1502,
      p1_name = "A",
      p2_name = "A",
      p3_name = "A"
    )
  
  multik_agri_plot$p3 
  multik_cons_plot$p3
```

10.2 Identify and classify edge types from the multi-kingdom network
```{r}
kingdom_vec = taxonomy_all$Domain
names(kingdom_vec) = taxonomy_all$Genus

cons_edges = get_edge_kingdoms(multiK_cons_net,kingdom_vec)
agri_edges = get_edge_kingdoms(multikK_agri_net,kingdom_vec)
```

10.3 Calculate the global proportions of edge types in the multi-kingdom networks
```{r}
cons_props = cons_edges |>
  dplyr::count(edge_class) |>
  dplyr::mutate(proportion = n/sum(n),management = "Conservation")

agri_props = agri_edges |>
  dplyr::count(edge_class) |>
  dplyr::mutate(proportion = n/sum(n),management = "Agriculture")

edge_props = rbind(cons_props, agri_props)
```

10.4 Plot global edge type composition
```{r}
agri_plot = plot_global_edge_composition(agri_props,"E")
cons_plot = plot_global_edge_composition(cons_props,"F")
```

10.5 Add modlue identity information to nodes in the kingdom vec
```{r}
agri_modules = get_module_composition(multiK_agri_modules$membership,kingdom_vec)
cons_modules = get_module_composition(multiK_cons_modules$membership,kingdom_vec)
```

10.6 Calculate the edge types within each module of the multi-kingdom network
```{r}
agri_module_edges = get_module_edge_types(multikK_agri_net,
                                          multiK_agri_modules$membership,
                                          kingdom_vec)

cons_module_edges = get_module_edge_types(multiK_cons_net,
                                          multiK_cons_modules$membership,
                                          kingdom_vec)
```

10.7 Preapre the data to generate a stream plot showing prportion of edge types within each module
```{r}
agri_stream = agri_module_edges |>
  dplyr::group_by(module) |>
  dplyr::mutate(prop = n / sum(n)) |>
  dplyr::ungroup()

cons_stream = cons_module_edges |>
  dplyr::group_by(module) |>
  dplyr::mutate(prop = n / sum(n)) |>
  dplyr::ungroup()

agri_stream = agri_stream |>
  dplyr::mutate(edge_class = factor(edge_class, levels = c("FF","BF","BB"))) |>
  dplyr::arrange(module, edge_class) |>
  dplyr::group_by(module) |>
  dplyr::mutate(ymin = cumsum(prop) - prop, ymax = cumsum(prop)) |>
  dplyr::ungroup()

cons_stream = cons_stream |>
  dplyr::mutate(edge_class = factor(edge_class, levels = c("FF","BF","BB"))) |>
  dplyr::arrange(module, edge_class) |>
  dplyr::group_by(module) |>
  dplyr::mutate(ymin = cumsum(prop) - prop, ymax = cumsum(prop)) |>
  dplyr::ungroup()

module_order = sort(unique(cons_module_edges$module))

stream_ribbons_cons = cons_stream |>
  dplyr::group_by(edge_class) |>
  dplyr::arrange(module) |>
  dplyr::mutate(x = seq_along(module)) |>
  dplyr::ungroup()

stream_ribbons_agri = agri_stream |>
  dplyr::group_by(edge_class) |>
  dplyr::arrange(module) |>
  dplyr::mutate(x = seq_along(module)) |>
  dplyr::ungroup()
```

10.8 Generate the stream plots
```{r}
agri_stream_plot = plot_module_streams(agri_stream,
                                       module_order = sort(unique(agri_stream$module)),
                                       plot_title = "C")

cons_stream_plot = plot_module_streams(cons_stream,
                                       module_order = sort(unique(cons_stream$module)),
                                       plot_title = "D")
```

10.9 Generation of a summary table of the proportion and weight of edge types in the multi-kingdom networks
```{r}
agri_signs_tab = get_edge_signs_by_class(multikK_agri_net,kingdom_vec)

cons_signs_tab = get_edge_signs_by_class(multiK_cons_net,kingdom_vec)
```
##############################


